\section{Introduction}

\subsection{Zsyntax}

The main aim of \cite{adding-logic} is to suggest that logic can play a
significant role in the toolbox of molecular biology, particularly in the
systematic formalization of certain molecular mechanisms characterized by
state transitions, that can therefore sensibly be described as ``inferential''.
The article shows how biochemical pathways, i.e., transitions from a molecular
aggregate to another molecular aggregate, can be viewed as deductive processes.

As already pointed out by Girard \cite{ll-syntax-semantics}, state transition
systems, like the ones considered in biochemistry, can be effectively described
by linear logic.  In essence, by representing chemical equations as axioms, the
notion of linear consequence corresponds to the notion of accessible state from
an initial one. Moreover, the complete description of the intermediate state
transitions can easily be extracted from the deduction itself.
In \cite{adding-logic}, the authors follow Girard's suggestion and outline a
correspondence between linear deductions and (bio)chemical reactions, observing,
however, that a modification is needed in order to satifactorily represent
the kind of state transitions that are typical of biomolecular models.

In order to represent the transition from a molecular aggregate to another
molecular aggregate as a deductive process, a logical system called Zsyntax is
used. The formal language of Zsyntax is inductively defined from a set of atomic
formulas (defined in \cite{adding-logic} as the bonding language
$\bioformulas$), combined by some linear logic-inspired operators:

\begin{enumerate}
\item aggregative Z-conjunction $\otimes$ (which corresponds to multiplicative
  conjunction)
\item Z-conditional $\rightarrow$ (which corresponds to linear implication)
\item selective Z-conjunction $\wedge$ (which corresponds to additive
  conjunction)
\item Z-disjunction $\vee$ (which corresponds to additive disjunction)
\item a unit formula $\top$ (which corresponds to the multiplicative conjunction
  unit)
\end{enumerate}

Given the meaning of the Zsyntax operators (thoroughly described in
\cite{adding-logic}), the formulas in the resulting Zsyntax language correspond
to suitable types of molecular aggregates. A proof of $\Gamma \models \Delta$
then represents the biochemical pathway from the aggregate $\Gamma$ to the
aggregate $\Delta$, where $\models$ is defined in \cite{adding-logic} as the
analogous of a logical consequence relation. Proofs are built according to a
logical calculus formulated in natural deduction style.

The main difference between Zsyntax and linear logic lies in the
non-monotonicity of the logical consequence relation (and therefore of the
linear implication operator, here called Z-conditional). To clarify,
monotonicity means that the validity of a formula $A \rightarrow B$ implies the
validity of $A \otimes C \rightarrow B \otimes C$ for any $C$. This is at odds
with the fact that many imiportant state transitions of interest in molecular
biology are context-sensitive: a reaction may not take place depending on the
molecular context in which it occurs.

The logical system described by the authors in therefore one that considers
context-sensitive state transitions: a transition may take place in every
molecular context that satisfies its control condition, which is a device used
to describe all contexts that instead inhibit such transition.
Control conditions are expressed in the formalism of Zsyntax as
\emph{control sets}, which are intended to be empirically determined (i.e., they
result from empirical knowledge obtained in the laboratory). The immediate
consequence is that their content typically changes over time, resulting in a
logical system that is \emph{open}: theorems may loose their status depending on
modification of the empirical knowledge.

\subsection{Automated deduction}

The objective of this report is to present the development of an automated
theorem prover for (a suitable fragment of) the calculus of Zsyntax presented in
\cite{adding-logic}. The work required a study of the proof-theoretic properties
of the logic, a survey of the already existing body of knowledge concerning
automated deduction for similar logics (and in particular linear logic), and the
adaptation of procedures from the literature to the specific case at hand.

The logic that has been implemented does not correspond to the full calculus
presented in \cite{adding-logic}, but it is a fragment that has been selected to
be sufficiently expressive to be practically useful, sufficiently representative
of the cornerstone concepts of Zsyntax (namely, controlled monotonicity and
openness of the system), and enough compact to be manageable in the short time
available. In retrospective, this choice proved to be quite appropriate, as the
sole study and implementation of the aforementioned fragment already took
several weeks to its completion.

The development presented here proceeded with a simple yet effective approach. A
new logical calculus for Zsyntax based on intuitionistic linear logic was
defined, with proof search in mind. In particular, this new logic was obtained
by first studying the properties that differentiate Zsyntax from plain linear
logic, and then adding them to a linear logic sequent calculus in the form of
annotations.  This yielded a calculus with derivations that are structurally
identical (modulo annotations) to the ones of linear logic, allowing us to (more
or less easily) exploit already existing automated deduction
procedures. Annotations could then be used as a way to tune, or restrict, the
inference engine to account for the biological constraints.

\subsection{Main references and related work}

The automated deduction parts of what follows are almost entirely based on the
work of K. Chaudhuri (\cite{chaudhuri-paper}, \cite{chaudhuri-thesis}). With his
work, collected in his PhD thesis, he shows how to exploit two well-known
automated reasoning techniques for linear logic, namely focusing and the inverse
method, in the same logical calculus, thus establishing a so called
\emph{focused inverse method} for linear logic.

Chaudhuri's work is targeted to the full intuitionistic linear logic, so it was
possible to adapt most of the material in \cite{chaudhuri-thesis}
straightforwardly, since the logic described here and implemented in the proof
assistant can be seen as a small (and much simpler) fragment of propositional
intuitionistic linear logic. The original parts presented here concern the
addition of control set-based facilities to the calculi of
\cite{chaudhuri-thesis}, and the modification of parts of them to limit the
amount of focusing, thus accounting for the order-sensitiveness of Zsyntax
deductions.

As already mentioned, the whole work was an attempt to implement (a suitable
part of) the logic presented in \cite{adding-logic}.
Other articles by the same authors were used as reference, in particular
\cite{non-mono} is where the idea of adding annotations to sequents comes from,
as well as the formalism used for control sets.
Contrary to \cite{adding-logic}, and therefore the work presented here,
\cite{non-mono} does not consider implication operators, but instead represents
empirical knowledge as additional axioms to be added to the formal system.
It also presents a more powerful conjunction operator, capable of expressing
when two aggregates are in the same linear context but do not interact with each
other.

\subsection{Outline}

Section 2 gives the reasons why plain Zsyntax as given in \cite{adding-logic} is
not entirely satisfactory when it comes to machine implementation, and then
proceeds by defining an alternative calculus that is very much related but
subtly different, and that sets the foundation for the rest of the work, as well
as the implemented software.

Section 3 gives a cut-free sequent calculus translation of the logic of the
previous section, and establishes soundness and completeness with respect to
it. A proof of cut admissibility is also given, which is required for the
completeness proof to go through.

Section 4 presents the idea of focusing as a way to improve proof search and
make it more computationally feasible by removing inessential
non-determinism. It then adapts the idea to our case, thus developing a focused
sequent calculus based on the standard one given in the previous section.

Section 5 presents the idea of forward reasoning as an alternative to the
more-frequently-used backward reasoning, and draws from the literature to show
how such alternative approach can be very effective to speed up inference for
linear logic and related formalisms, like the one used here.  It then again
takes [cmu thesis] as a model to represent the focused calculus of the preceding
section in a forward-oriented direction.

Section 6 explains the benefits of the forward focused calculus developed in the
previous sections, and gives some details on how to go from the definition of
the calculus to an implementation of it.

Section 7 gives some implementation details on the search procedure, based on
the forward focused calculus.

Section 8 ...

Section 9 gives a notion of derivation terms, that is first class objects that
can be used to represent derivations in the sequent calculus in a compact way
that is particularly useful in implementations.

Section 10 concludes with some remarks on various ways in which the work
presented here could be extended in the future.










%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../docs"
%%% End:
